---
title: "Exploratory Analysis"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Distribution of Values

```{r, echo=FALSE, include=FALSE, message=FALSE}
source("map.R")
source("causes.R")
```

## Statistical Analysis
The data that we’re most interested in is data about the percentage of a regions population that is using unimproved drinking water. 

```{r , echo=FALSE, include=FALSE}
summary(poor_water$unimproved_dwater)
```

Analyzing the data collected from UNICEF we can see that the mean proportion of people using unimproved drinking water sources is rather low, at 4.85%. Although this value is low it’s important to note that 25% of countries have between 5.85 % of their population and 38.72% of their population using unimproved water sources. This is likely because despite most of the world having very low usage of unimproved water sources, there still exist countries with very high usage of unimproved water sources. This becomes more obvious when we note that despite the median being 1.1% the mean percentage of people using unimproved water is 5.85%. 

At further inspection we can see that the standard deviation of the proportion of populations using unimproved drinking water is 7.513%. This supports our analysis that there exist countries which use unimproved water at significantly higher rates than the other regions in the world. 

## Geographic Distribution 

Unimproved drinking water sources include unprotected dug wells, unprotected springs, cart with small tank/drum, bottled water, tanker truck, and surface water (river, dam, lake, pond, stream, canal, irrigation channels). The data gives us significant insights on which parts of the world might be potentially affected in terms of health. We notice that the African continent has significantly higher proportions of populations using unimproved sources of drinking water. Causes of health burden across the world measured against the data on proportions of population using unimproved water sources can bring to light the countries that are most affected by this particular indicator and would thus help in better development of strategies for devicing solutions (starting from improving water sources).

```{r map, echo=FALSE, message=FALSE}
mapParams <- mapCountryData( water_data_map_joined
                            , nameColumnToPlot="Proportion Using Unimproved Sources of Drinking Water"
                            , addLegend=FALSE
                            , catMethod = catMethod
                            , colourPalette = colourPalette )
do.call( addMapLegend , c( mapParams
                          , legendLabels="all"
                          , legendWidth=0.5
                          , legendIntervals="data" , legendMar = 2 ))
```

## Common Causes of Health Burden 

The plot below is a histogram of the most frequent diseases found in the bottom 10% of countries with the lowest access to clean drinking water. Countries were ranked based on the proportion of their population that is reliant on an unimproved water source. For each of the countries within the bottom 10%, the top 10 causes which contributed the most to their DALYs were selected and aggregated to create a count of significant causes of health burden among countries with low access to clean drinking water. 
```{r, echo=FALSE, message=FALSE}
plot(common_causes)
```

### Interpretation
From the histogram we learn that the most significant cause of health burden among regions without access to clean water are diseases relating to newborn children. Other frequent causes of health burden include STDs, malaria, and nutritional diseases. Identifying the frequent causes of health burden can help us understand how poor access to clean drinking water greatly affects the lives of children, and parents. This makes reproduction difficult for people in such countries, because they’re risk of health burden is much higher.  Countries with such problems may have trouble in the future, if they’re is difficult to keep young children healthy. It is important to assist such countries in acquiring better access to clean water so that their populations can more independently support themselves in the future. 

#Relationships Between Variables 
# Relationships of DALYs rate and water quality of top 8 DALYS rate countries

The purpose of these plots is to examine how high DALYs rate countries access to limited water sources, and unimproved water sources in 2017. This can recognize which country has both limited and unimproved water sources or which country has limited or unimproved water sources. First, I find top 10 countries with high DALYs rate, and compare their proportion of population of access to limited water sources. It has two NA variables for water sources which I delete those columns for a better scatterplot. I also did the same interpretation for proportion of population of access to unimproved water sources. These two variables are important to which countries have poor access to drinking water such as unimproved and limited sources related to high DALYs rate.
```{r plot, echo=FALSE, message=FALSE}
source("dalys_water.R")
plot(dalys_limitedwater)

plot(dalys_unimproved)
```

Plot of DALYs rate versus limited drinking water sources have seems to have many African countries which we assume they have limited drinking water sources and high DALYs rate. The most impacted country by both DALYs rate and proportion of population access to limited water sources is Somalia which have quite high prop of limited drinking water and at top three of DALYs rate countries. 
Next plot is DALYs rate versus unimproved drinking water source also shows many African countries which we assume they have unimproved drinking water sources and high DALYs rate. The most impacted country by both DALYs rate and proportion of population access to unimproved water sources is Mali which has quite high prop of unimproved drinking water source and highest DALYs rate of top 8 countries.
In both plots, Somalia has both high impacted on limited drinking water sources, and unimproved water sources which it have high rate of DALYs as well. 



```{r, warning=FALSE, message=FALSE, include=FALSE}
ihme <- read.csv('IHME.csv', header = T)
water <- read.csv('water2.csv', header = T)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r data prep, include=FALSE}
Diarrhea <- ihme %>% 
  filter(cause == "Diarrheal diseases" & metric == "Rate")
Intestinal <- ihme %>% 
  filter(cause == "Intestinal nematode infections" & metric == "Rate")
Maternal <- ihme %>% 
  filter(cause == "Maternal sepsis and other maternal infections" & metric == "Rate")
Deficiencies <- ihme %>% 
  filter(cause == "Nutritional deficiencies" & metric == "Rate")

names(water)[names(water) == "Geographic.Area"] <- "location"
wd <- merge(Diarrhea, water, by = "location")
wi <- merge(Intestinal, water, by = "location")
wm <- merge(Maternal, water, by = "location")
wdef <- merge(Deficiencies, water, by = "location")

wd_dalys <- wd %>% 
  filter(measure == "DALYs (Disability-Adjusted Life Years)")
wi_dalys <- wi %>% 
  filter(measure == "DALYs (Disability-Adjusted Life Years)")
wm_dalys <- wm %>% 
  filter(measure == "DALYs (Disability-Adjusted Life Years)")
wdef_dalys <- wdef %>% 
  filter(measure == "DALYs (Disability-Adjusted Life Years)")
```

## Plots of relationship between various causes & water access measured by DALYS

The purpose of these plots is to explore if there is a correlation between
limited drinking water access and diseases or infections that may be associated
with that. If there is a relationship then we can look into causal effects and 
this provides a good visual for how strong that relationship may be.
In this dataset, **limited drinking water access is defined as water collection 
from an improved source exceeded 30 minutes**. Some of the causes we looked at 
were based on background information regarding water collection, such as
women and girls bear the majority of the burden of water collection so we wanted
to examine the relationship between maternal sepsis and other maternal 
infections.

```{r plot1, warning = FALSE, echo=FALSE, message=FALSE}
ggplot(wd_dalys, aes(x= val, y=X2017_Proportion.of.population.using.limited.drinking.water.services)) + 
  geom_point()+ labs(title = "Rate of Diarrheal diseases vs. 
Proportion of pop using limited drinking water",
x = "DALYs due to Diarrheal Diseases", y = " Proportion of Population using 
limited drinking water")+
  geom_smooth(method="loess", se=F)
```

This plot shows the relationship between diarrheal disease and proportion of the
population using limited drinking water. There is a trend upward, as the proportion
of the population with limited drinking water increases, DALYs due to diarrheal
diseases also increases with a few outliers. This is expected becauase some 
diarrheal diseases are caused by poor water quality.

```{r plot2, warning = FALSE, echo=FALSE, message=FALSE}
ggplot(wi_dalys, aes(x= val, y=X2017_Proportion.of.population.using.limited.drinking.water.services)) + 
  geom_point()+ labs(title = "Rate of Intestinal nematode infections vs. 
Proportion of pop using limited drinking water", x = "DALYs due to Intestinal nematode infections", y = " Proportion of Population using limited drinking water")+
  geom_smooth(method="loess", se=F)
```

The plot above shows the relationship between intestinal nematode infections
and access to drinking water. The graph shows an upward trend, similar to 
diarrheal diseases. The greater percentage of the population with limited access
to drinking water, the higher the DALYs due to nematode infections. 

```{r plot3, warning = FALSE, echo=FALSE, message=FALSE}
ggplot(wm_dalys, aes(x= val, y=X2017_Proportion.of.population.using.limited.drinking.water.services)) + 
  geom_point()+
  geom_smooth(method="loess", se=F)+
  labs(title = "Rate of Maternal sepsis & other maternal infections vs. 
Proportion of pop using limited drinking water", 
x = "DALYs due Maternal sepsis and other maternal infections", y = " Proportion of Population using limited drinking water")
```

The plot above shows the relationship between maternal sepsis and other maternal
infections and access to drinking water. As mentioned before, the burden of 
getting water falls unevenly on women and girls, so we decided to explore this
area. The trend line shows an increase and then it plateaus, which is indicative
of an unclear relationship. The points are somewhat scattered compared to the 
two previous causes and call for further investigation.

```{r plot4, warning = FALSE, echo=FALSE, message=FALSE}
ggplot(wdef_dalys, aes(x= val, y=X2017_Proportion.of.population.using.limited.drinking.water.services)) + 
  geom_point()+ geom_smooth(method="loess", se=F)+
  labs(title = "Rate of Nutritional deficiencies vs. Proportion of pop using limited 
drinking water", x = "DALYs due Maternal sepsis and other maternal infections", 
y = " Proportion of Population using limited drinking water")
```

The plot above shows the relationship between nutritional deficiencies and 
access to drinking water. This graph follows a similar pattern to the maternal 
sepsis and other maternal infections graph with an increase and then a dropoff,
which is likely caused by overfitting to an outlier (Mali). In Mali, there is 
one clear point which has a high rate of maternal sepsis and infection but 
access to limited drinking water is not as high as the rest of the plot would
predict. This goes to show that these plots represent correlation much more than
they do causation, and any insights gathered require more analysis to be able
to accurately make statements regarding them.